Appendix: Why the complement of AUC isn't a casewise CV criterion

### 2024-07-04 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
incorporate new text in the JSS article for the section on mixed models from the revised vignette.

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
palette(car::carPalette())
```

@@

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error. The $n$ regression models are highly statistically dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has larger variance than if the predictions from the models fit to the $n$ data sets were independent.

Because predictions are based on smaller data sets, each of size approximately $n - n/k$, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) is more biased than estimated prediction error for LOO CV. It is possible, however, to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

The relative *variance* of prediction error for LOO CV and $k$-fold CV (with $k < n$) is more complicated: Because the overlap between the data sets with each fold omitted is smaller for $k$-fold CV than for LOO CV, the dependencies among the predictions are smaller for the former than for the latter, tending to produce smaller variance in prediction error for $k$-fold CV. In contrast, there are factors that tend to inflate the variance of prediction error in $k$-fold CV, including the reduced size of the data sets with each fold omitted and the randomness induced by the selection of folds---in LOO CV the folds are not random.

@@

In this section, we introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models. Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model. Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.

Consider a researcher studying improvement in a skill, singing, for example, among students enrolled in a four-year voice program at a music conservatory. The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in 5 annual measurements for each student. It turns out that singing appeals to students of all ages, and students enrolling in the program range in age from 20 to 70. Moreover, participants' untrained singing skill is similar at all ages, as is their rate of progress with training. All students complete the four-year program.

The researcher, who has more expertise in singing than in modeling, decides to model the response, $y$, singing skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time. The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.

We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]

[^1]: We invite the interested reader to experiment with varying the parameters of our example.

```{r data}
# Parameters:
set.seed(9693) 
Nb <- 100     # number of groups
Nw <- 5       # number of individuals within groups
Bb <- 0       # between-group regression coefficient on group mean
SDre <- 2.0   # between-group SD of random level relative to group mean of x
SDwithin <- 0.5  # within group SD
Bw <- 1          # within group effect of x
Ay <- 10         # intercept for response
Ax <- 20         # starting level of x
Nx <- Nw*10      # number of distinct x values

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = Ax + rep(1:Nx, length.out = Nw*Nb)
) |> within ({
  xm  <- ave(x, group, FUN = mean) # within-group mean
  y <- Ay +
    Bb * xm +                    # contextual effect
    Bw * (x - xm) +              # within-group effect
    rnorm(Nb, sd=SDre)[group] +  # random level by group
    rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
})
```

Figure \ref{fig:plot1} (a) shows a scatterplot the data for a representative group of 10 (without loss of generality, the first 10) of the 100 students, displaying the 95% concentration ellipse for each cluster.

```{r plot1, echo=FALSE, results=FALSE}
#| fig.height = 4,
#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data",
#| fig.subcap=rep("", 5),
#| fig.ncol = 2, out.width = "45%", fig.align = "center"
library("lattice")
library("latticeExtra")
library("lme4", quietly=TRUE)
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot # display graph

mod.0 <- lmer(y ~ 1 + (1 | group), Data)
mod.1 <- lmer(y ~ x + (1 | group), Data)
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)

Data <- within(Data, {
  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
  fit_mod1.re <- predict(mod.1)
  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
  fit_mod2.re <- predict(mod.2)
  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
  fit_mod3.re <- predict(mod.3)
})

Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))
Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 2), lwd=2)))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35) #,
  # key=list(
  #   corner=c(0.95, 0.05),
  #   text=list(c("fixed effects only","fixed and random")),
  #   lines=list(lty=c(1, 3)))
  )

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16) #,
  # key=list(
  #   corner=c(0.05, 0.05),
  #   text=list(c("fixed effects only","fixed and random")),
  #   lines=list(lty=c(1, 3)))
  )

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16) #,
  # key=list(
  #   corner=c(0.05, 0.05),
  #   text=list(c("fixed effects only","fixed and random")),
  #   lines=list(lty=c(1, 3)))
  )

```

The between-student effect of age is 0 but the within-student effect is 1. Due to the large variation in ages between students, the least-squares regression of singing skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:

```{r}
summary(lm(y ~ x, data=Data))
```

We proceed to fit several mixed-effects models to the data, using the `compareCoefs()` function from the \pkg{car} package [@FoxWeisberg:2019] to display the fixed-effect estimates for these models. We discuss each of these models below.
```{r fitted-mixed-models}
mod.0 <- lmer(y ~ 1 + (1 | group), Data)
mod.1 <- lmer(y ~ x + (1 | group), Data)
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
library("car", quietly=TRUE)
compareCoefs(mod.0, mod.1, mod.2, mod.3)
```

The initial mixed-effects model, \code{mod.0}, is a simple random-intercepts model. We obtain predictions from this model for the fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students). Predictions from \code{mod.0} for the first 10 students are shown in Figure \ref{fig:plot1} (b). The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean. Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.

Our next model, \code{mod.1}, includes a fixed intercept and the fixed effect of \code{x}, along with a random intercept. Predictions from this model appear in Figure \ref{fig:plot1} (c). The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}. We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod.1} to \code{mod.0}, but that cluster-based cross-validation will prefer \code{mod.0} to \code{mod.1}.

Our third model, \code{mod.2}, includes the "contextual effect" of $x$---that is, the cluster mean \code{xm}---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept. This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).

Predictions from \code{mod.2} appear in Figure \ref{fig:plot1} (d). Depending on the estimated variance parameters of the model, a mixed model like \code{mod.2} will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students. In our contrived data, \code{mod.2} applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student. The fit of the mixed model \code{mod.2} is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).

The  mixed model \code{mod.2} therefore fits the individual observations well, and we anticipate a favorable assessment using individual-based cross-validation. In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for  \code{mod.2} compared to the smaller \code{mod.0}, which includes only fixed and random intercepts.

Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model. In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.

Finally, \code{mod.3} directly estimates the model used to generate the data. As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}. The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar to those from \code{mod.2}.

We next carry out case-based cross-validation, which, as we have explained, includes both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which includes fixed effects only. In order to reduce between-model random variability in comparisons of models, we apply \code{cv()} to the list of models created by the \code{models()} function (introduced previously), performing cross-validation with the same folds for each model (see Figure \ref{fig:cross-validation-clusters}):

```{r cross-validation-clusters}
#| fig.height = 5,
#| fig.subcap=rep("", 2),
#| fig.ncol = 2, out.width = "45%", fig.align = "center",
#| fig.cap = "10-fold (a) cluster-based and (b) case-based cross-validation comparing random intercept models with varying fixed effects. The error bars show the 95% confidence interval around the CV estimate of the MSE for each model."
modlist <- models("~ 1"=mod.0, "~ 1 + x"=mod.1, 
                  "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)

cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")

cvs_cases <- cv(modlist, data=Data, seed=9693)
plot(cvs_cases, main="Model Comparison, Case-Based CV")
```

In summary, \code{mod.1}, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. \code{mod.2}, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV. The data-generating model, \code{mod.3}, which includes the fixed effect of \code{x - xm} in place of separate terms in \code{x} and \code{xm}, isn't distinguishable from \code{mod.2}, which includes \code{x} and \code{xm} separately, even though \code{mod.2} has an unnecessary parameter (recall that the population coefficient of \code{xm} is 0 when \code{x} is expressed as deviations from the contextual mean). These conclusions are consistent with our observations based on graphing predictions from the various models in Figure \ref{fig:plot1} (on page \pageref{fig:plot1}), and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

>>>>

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
palette(car::carPalette())
library(car)     # included this to avoid scatterplotMatrix() error
```

@@

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error. The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance. In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance. It is also possible to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

@@

We introduce four artificial data sets that exemplify aspects of cross-validation particular to hierarchical models. Using these data sets, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model. Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.

The four datasets differ in the relative magnitude of between-cluster variance compared with within-cluster variance. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.

For these analyses, we use the `glmmTMB()` function in the **glmmTMB** package [@BrooksEtAl] because, in our experience, it is more likely to converge than functions in the **nlme** and the **lme4** packages for models with low between-cluster variance.

Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease. The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.  The data are observational, with dosages prescribed by the patients' physicians, so that patients who were more severely affected by the disease received higher dosages of the drug.

Our four contrived data sets (see below) illustrate possible results for data obtained in such a scenario. The relative configuration of dosages `x` and symptoms `y` are identical within patients in each of the four datasets. Within patients, higher dosages are generally associated with a reduction in symptoms.

Between patients, however, higher dosages are associated with higher levels of symptoms. A plausible mechanism is a reversal of causality: Within patients, higher dosages alleviate symptoms, but, between patients, higher morbidity causes the prescription of higher dosages.

The four data sets differ in the between-patient variance of patient centroids from a common between-patients regression line. The data sets exhibit a progression from low to high variance around the common regression line.

We start by generating a data set to serve as a common template for the four sample data sets.  We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template]

[^data-template]: We invite the interested reader to experiment with varying the parameters of our example.

```{r include=FALSE, echo=FALSE}
library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r parameters}
# Parameters:
Nb <- 20     # number of patients
Nw <- 5      # number of occasions for each patient
Bb <- 1.0    # between-patient regression coefficient on patient means
Bw <- -0.5   # within-patient effect of x

SD_between <- c(0, 5, 6, 8)               # SD between patients
SD_within <- rep(2.5, length(SD_between)) # SD within patients

Nv <- length(SD_within)       # number of variance profiles
SD_ratio <- paste0('SD ratio = ', SD_between,' / ',SD_within)
SD_ratio <- factor(SD_ratio, levels = SD_ratio)

set.seed(833885)

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2, 2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean
    # Scaled random error within each SD_ratio_i group
    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })
Data <- do.call(
  rbind,
  lapply(
    1:Nv,
    function(i) {
      cbind(Data_template, SD_ratio_i = i)
    }
  )
)
Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```
Figure \ref{fig:plot1} display scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]

[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1, echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       par.strip.text = list(cex = 0.7),
       scales = list(x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```
The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$. The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four datasets for a total of 16 models.

```{r model-fits}
model.formulas <- c(
  ' ~ 1'             =  y ~ 1 + (1 | patient),
  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)
fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```
We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r predict}
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
pred.BLUPs <- lapply(fits, lapply, predict)
```
<!-- We then prepare the data for plotting: -->

```{r data-predictions,include=FALSE}
Dataf <- lapply(split(Data, ~ SD_ratio),
                function(d) {
                  lapply(names(model.formulas), 
                         function(form) cbind(d, formula = form))
                }) |> 
  lapply(function(dlist) do.call(rbind, dlist)) |> 
  do.call(rbind, args = _) |> 
  within(
    {
      pred.fixed <- unlist(pred.fixed)
      pred.BLUPs <- unlist(pred.BLUPs)
      panel <- factor(formula, levels = c(names(model.formulas), 'data'))
    }
  )
Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
```
The fixed-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-fixed} along with a summary of the data using 50% concentration ellipses.  Figure \ref{fig:plot-fits-blups} similarly displays predicted BLUPs. 
```{r plot-fits-fixed,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed-effect predictions using each model applied to data sets with varying between- and within-patient variance ratios.  The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
{
  xyplot(y ~ x |SD_ratio * panel, Data,
         groups = patient, type = 'n',    
         par.strip.text = list(cex = 0.6),
         scales = list(x = list(alternating = FALSE)),
         drop.unused.levels = FALSE,
         ylab = list('fixed-effect predictions',cex=.7)) + 
    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                         level = 0.5),
           panel.abline(a = 10, b = 1)) +
    xyplot(pred.fixed  ~ x |SD_ratio * panel, Dataf, type = 'l', 
           groups = patient,
           drop.unused.levels = F)
}|> useOuterStrips()
```
<!-- The BLUPs from these models, shown in Figure \ref{fig:plot-fits-blups}, are produced as follows: -->
```{r plot-fits-blups,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios. The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
{   xyplot(y ~ x | SD_ratio * panel, Data,
         groups = patient, type = 'n',
         drop.unused.levels = F, 
         par.strip.text = list(cex = 0.6),
         scales = list(x = list(alternating = FALSE)),
         ylab = list('fixed- and random-effect predictions (BLUPS)', cex = 0.7)) +
    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                         level = 0.5),
           panel.abline(a = 10, b = 1)) +
    xyplot(pred.BLUPs  ~ x | SD_ratio * panel, Dataf, type = 'l', 
           groups = patient,
           drop.unused.levels = F) 
}|> useOuterStrips()
```
Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two datasets with the smallest between-patient variances.

For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for datasets with larger between-subject variance.  With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, `~ 1`  to the larger `~ 1 + x` model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former.

We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' dataset, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the '8 / 2.5' dataset.

The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, along with a random intercept. This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.  The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only. In order to reduce between-model random variability in comparisons of models on the same dataset, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

```{r echo=FALSE,include=FALSE}
library(cv) # unclear why it's necessary to reload cv
.opts <- options(warn = -2) 
```
```{r cross-validation,cache=TRUE,message=FALSE}
model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
cvs_cases <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]], k = 10, 
              data = split(Data, ~ SD_ratio)[[i]])
         })
cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, 
              data = split(Data, ~SD_ratio)[[i]], 
              clusterVariables = 'patient')
         })
```
For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. For example, this code produces Figure \ref{fig:plot-cv-example}.
```{r plot-cv-example}
#| out.width = "60%",
#| fig.height = 3.5,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
plot(cvs_clusters[[2]], main="Comparison of Fixed Effects",)
```
We assemble the results for all datasets to show them in a common figure.
```{r cross-validation-data}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', reverse.rows = T, columns = 2), 
       type = 'b', ylab = 'CV criterion (MSE)',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 2),
              superpose.symbol=list(pch = 15:16, cex = 1)),
       scales = list(y = list(log = TRUE), 
                     x = list(alternating = F, rot = 45))) 
```
In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
